Online Multiclass Linear Classification with Bandit Feedback

### 2019-01-18 ICML-submission/from-weak-separability-to-strong-separability.tex
Fixes

According to the well known Stone-Weierstrass theorem~\citep[see
e.g.][Section~10.10]{Davidson-Donsig-2010}, on a compact set,
multivariate polynomials uniformly approximate any continuous function.
Intuitively speaking, we use a multivariate polynomial to approximate, on the
unit ball of $\R^d$, some scaling of the indicator function corresponding to the intersection of
$m=K-1$ halfspaces $\bigcap_{j \in \cbr{1,\ldots,K} \setminus \cbr{i}} \cbr{x: \ip{w_i^* - w_j^*}{x} \geq 0}$. Within margin $\gamma$ along the decision boundary, we allow
the polynomial to attain arbitrary value. To be able to quantify $\gamma'$
we need to quantify certain parameters of the approximating polynomial.
Following the constructions in~\citet{Klivans-Servedio-2008}
(which in turn use the constructions in~\citet{Beigel-Reingold-Spielman-1995}), we
construct two different polynomials with different parameters that satisfy
Equation~\eqref{eqn:r-plus} and~\eqref{eqn:r-minus}.
The parameters
are quantified in
Theorems~\ref{theorem:polynomial-approximation-1}~and~\ref{theorem:polynomial-approximation-2}
stated below.

%The polynomial separates examples in
%one class from examples in the other classes.
%
%the indicator function corresponding to the intersection of
%$m=K-1$ halfspaces
%a function that takes value $\frac {\gamma'} 2$ on $R_i^+$
%and $-\frac {\gamma'} 2$ on $R_i^-$

\paragraph{Details of the proof.} Before we state the theorems, recall that a polynomial of $d$ variables is a
function $p:\R^d \to \R$ of the form
\begin{align*}
p(x)
& = p(x_1, x_2, \dots, x_d) \\
& = \sum_{\alpha_1, \alpha_2, \dots, \alpha_d} c_{\alpha_1, \alpha_2, \dots, \alpha_d} x_1^{\alpha_1} x_2^{\alpha_2} \dots x_d^{\alpha_d}
\end{align*}
where the sum ranges over a finite set of $d$-tuples $(\alpha_1, \alpha_2,
\dots, \alpha_d)$ of non-negative integers and $c_{\alpha_1, \alpha_2, \dots,
\alpha_d}$'s are real coefficients. The \emph{degree} of a polynomial $p$,
denoted by $\deg(p)$, is the largest value of $\alpha_1 + \alpha_2 + \dots +
\alpha_d$ for which the coefficient $c_{\alpha_1, \alpha_2, \dots, \alpha_d}$ is
non-zero. Following the terminology of~\citet{Klivans-Servedio-2008}, the
\emph{norm of a polynomial} $p$ is defined as
$$
\norm{p} = \sqrt{\sum_{\alpha_1, \alpha_2, \dots, \alpha_d} \left(c_{\alpha_1, \alpha_2, \dots, \alpha_d} \right)^2 } \; .
$$
It is easy see that this is indeed a norm, since we can interpret it as the
Euclidean norm of the vector of the coefficients of the polynomial.

\begin{theorem}[Polynomial approximation of intersection of halfspaces I]
\label{theorem:polynomial-approximation-1}
Let $v_1, v_2, \dots, v_m \in \R^d$ be vectors such that $\norm{v_1},
\norm{v_2}, \dots, \norm{v_m} \le 1$. Let $\gamma \in (0,1)$. There exists a
multivariate polynomial $p:\R^d \to \R$ such that
\begin{enumerate}
\item $p(x) \ge 1/2$ \\ for all $\displaystyle x \in \bigcap_{i=1}^m \left\{ x \in \R^d ~:~ \norm{x} \le 1, \ \ip{v_i}{x} \ge \gamma \right\}$,
\item $p(x) \le -1/2$ \\ for all $\displaystyle x \in \bigcup_{i=1}^m \left\{ x \in \R^d ~:~ \norm{x} \le 1, \ \ip{v_i}{x} \le - \gamma \right\}$,
\item $\displaystyle \deg(p) = \left\lceil \log_2(2m) \right\rceil \cdot \left\lceil \sqrt{\frac{1}{\gamma}} \right\rceil$,
\item $\displaystyle \norm{p} \le \left( 188 \left\lceil \log_2(2m) \right\rceil \cdot \left\lceil \sqrt{\frac{1}{\gamma}} \right\rceil \right)^{\frac{1}{2} \left\lceil \log_2(2m) \right\rceil \cdot \left\lceil \sqrt{\frac{1}{\gamma}} \right\rceil}$.
\end{enumerate}
\end{theorem}

\begin{theorem}[Polynomial approximation of intersection of halfspaces II]
\label{theorem:polynomial-approximation-2}
Let $v_1, v_2, \dots, v_m \in \R^d$ be vectors such that $\norm{v_1},
\norm{v_2}, \dots, \norm{v_m} \le 1$. Let $\gamma \in (0,1)$.
Define
$$
r = 2 \left\lceil \frac{1}{4} \log_2(4m + 1) \right\rceil + 1 \quad \text{and} \quad s = \left \lceil \log_2(1/\gamma) \right \rceil \; .
$$
Then, there exists a multivariate polynomial $p:\R^d \to \R$ such that
\begin{enumerate}
\item $\displaystyle p(x) \ge 1/2$ \\
for all $\displaystyle x \in \bigcap_{i=1}^m \left\{ x \in \R^d ~:~ \norm{x} \le 1, \ \ip{v_i}{x} \ge \gamma \right\}$,

\item $\displaystyle p(x) \le - 1/2$ \\
for all $\displaystyle x \in \bigcup_{i=1}^m \left\{ x \in \R^d ~:~ \norm{x} \le 1, \ \ip{v_i}{x} \le - \gamma \right\}$,

@@

The proofs of the theorems can be found in
Section~\ref{section:proof-of-polynomial-approximation}. The geometric
interpretation of the two regions described in parts 1 and 2 of the theorems is
explained in Figure~\ref{figure:pizza-slice}. Similar but weaker results were
proved by~\citet{Klivans-Servedio-2008}. In particular, our bounds in parts 1,
2, 3, 4 of
Theorems~\ref{theorem:polynomial-approximation-1}~and~\ref{theorem:polynomial-approximation-2}
are independent of the dimension $d$.

\begin{figure}
\begin{center}
\input{figures/pizza-slice}
\end{center}
\caption[]{The figure shows the two regions used in parts 1 and 2 of
Theorems~\ref{theorem:polynomial-approximation-1}~and~\ref{theorem:polynomial-approximation-2}
for the case $m=d=2$ and a particular choice of vectors $v_1, v_2$ and margin
parameter $\gamma$. These regions are
$R^+ = \displaystyle \bigcap_{i=1}^m \left\{ x \in \R^d ~:~ \norm{x} \le 1, \ \ip{v_i}{x} \ge \gamma \right\}$,
$R^- = \displaystyle \bigcup_{i=1}^m \left\{ x \in \R^d ~:~ \norm{x} \le 1, \ \ip{v_i}{x} \le - \gamma \right\}$
The separating hyperplanes $\ip{v_1}{x} = 0$ and $\ip{v_2}{x} = 0$ are shown as dashed lines.}
\label{figure:pizza-slice}
\end{figure}

@@

Using the lemma and the polynomial approximation theorems, we can prove the
mapping $\phi$ maps any set of weakly linearly separable examples
to a strongly linearly separable set of examples. Due to space constraints,
we only sketch the
proof of Theorem~\ref{theorem:margin-transformation} here, and defer the full
proof to Appendix~\ref{section:full-proof-of-theorem-margin-transformation}.

\begin{proof}[Proof sketch of Theorem~\ref{theorem:margin-transformation}]
Since the examples $(x_1, y_1)$, $(x_2, y_2)$, $\dots$, $(x_T, y_T)$ are weakly
linearly separable with margin $\gamma$,, there are vectors $w_1, w_2, \dots, w_K$
satisfying \eqref{equation:weak-linear-separability-1} and
\eqref{equation:weak-linear-separability-2}.

@@

\begin{align*}
\deg(p_i) & = \lceil \log_2(2K-2) \rceil \cdot \left\lceil \sqrt{\frac{2}{\gamma}} \right\rceil \; ,
\end{align*}
such that all examples $x$ in $R_i^+$ (resp. $R_i^-$) satisfy $p_i(x) \geq 1/2$
(resp. $p_i(x) \leq -1/2$).
Therefore, for all $t=1,2,\dots,T$, if $y_t = i$ then $p_i(x_t) \ge 1/2$,
 and if $y_t \neq i$ then $p_i(x_t) \le -1/2$, and
\begin{align*}
\norm{p_i} & \le \left(188 \lceil \log_2(2K-2) \rceil \cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil \right)^{\frac{1}{2} \lceil \log_2(2K-2) \rceil
\cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil} \; , \\
\end{align*}
By \autoref{lemma:norm-bound}, there exists $c_i \in \ell_2$ such that
$\ip{c_i}{\phi(x)} = p_i(x)$, and
\begin{align*}
\norm{c_i}_{\ell_2}
& \le \left(376 \lceil \log_2(2K-2) \rceil \cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil \right)^{\frac{1}{2} \lceil \log_2(2K-2) \rceil
\cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil} \; .
\end{align*}
Define vectors $u_i \in \ell_2$ as
\begin{align*}
u_i & = \frac{c_i}{\sqrt{K} \left(376 \lceil \log_2(2K-2) \rceil \cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil \right)^{\frac{1}{2} \lceil \log_2(2K-2) \rceil
\cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil}} \; . \\
\end{align*}

>>>>

According to the well known Stone-Weierstrass theorem~\citep[see
e.g.][Section~10.10]{Davidson-Donsig-2010}, on a compact set, multivariate
polynomials uniformly approximate any continuous function. Roughly speaking, the
conditions \eqref{eqn:r-plus} and \eqref{eqn:r-minus} mean that $p_i$
approximates on $\B(0,1)$ a scalar multiple of the indicator function of the
intersection of $K-1$ halfspaces $\bigcap_{j \in \cbr{1,2,\dots,K} \setminus
\cbr{i}} \cbr{x: \ip{w_i^* - w_j^*}{x} \geq 0}$ while within margin $\gamma$ along
the decision boundary, the polynomial is allowed to attain arbitrary values.
It is thus clear such a polynomial exists.

We give two explicit constructions for such polynomial in
Theorems~\ref{theorem:polynomial-approximation-1}~and~\ref{theorem:polynomial-approximation-2}.
Our constructions are based on \citet{Klivans-Servedio-2008} which in turn uses
the constructions from~\citet{Beigel-Reingold-Spielman-1995}. More importantly,
the theorems quantify certain parameters of the polynomial, which will allow us
to give an upper bound on the transformed margin $\gamma'$.

Before we state the theorems, recall that a polynomial of $d$ variables is a
function $p:\R^d \to \R$ of the form
\begin{align*}
p(x)
& = p(x_1, x_2, \dots, x_d) \\
& = \sum_{\alpha_1, \alpha_2, \dots, \alpha_d} c_{\alpha_1, \alpha_2, \dots, \alpha_d} x_1^{\alpha_1} x_2^{\alpha_2} \dots x_d^{\alpha_d}
\end{align*}
where the sum ranges over a finite set of $d$-tuples $(\alpha_1, \alpha_2,
\dots, \alpha_d)$ of non-negative integers and $c_{\alpha_1, \alpha_2, \dots,
\alpha_d}$'s are real coefficients. The \emph{degree} of a polynomial $p$,
denoted by $\deg(p)$, is the largest value of $\alpha_1 + \alpha_2 + \dots +
\alpha_d$ for which the coefficient $c_{\alpha_1, \alpha_2, \dots, \alpha_d}$ is
non-zero. Following the terminology of~\citet{Klivans-Servedio-2008}, the
\emph{norm of a polynomial} $p$ is defined as
$$
\norm{p} = \sqrt{\sum_{\alpha_1, \alpha_2, \dots, \alpha_d} \left(c_{\alpha_1, \alpha_2, \dots, \alpha_d} \right)^2 } \; .
$$
It is easy see that this is indeed a norm, since we can interpret it as the
Euclidean norm of the vector of the coefficients of the polynomial.

\begin{theorem}[Polynomial approximation of intersection of halfspaces I]
\label{theorem:polynomial-approximation-1}
Let $v_1, v_2, \dots, v_m \in \R^d$ be vectors such that $\norm{v_1},
\norm{v_2}, \dots, \norm{v_m} \le 1$. Let $\gamma \in (0,1)$. There exists a
multivariate polynomial $p:\R^d \to \R$ such that
\begin{enumerate}
\item $p(x) \ge 1/2$ \\ for all $\displaystyle x \in R^+ = \bigcap_{i=1}^m \left\{ x \in \B(0,1) ~:~  \ip{v_i}{x} \ge \gamma \right\}$,
\item $p(x) \le -1/2$ \\ for all $\displaystyle x \in R^- = \bigcup_{i=1}^m \left\{ x \in \B(0,1) ~:~  \ip{v_i}{x} \le - \gamma \right\}$,
\item $\displaystyle \deg(p) = \left\lceil \log_2(2m) \right\rceil \cdot \left\lceil \sqrt{\frac{1}{\gamma}} \right\rceil$,
\item $\displaystyle \norm{p} \le \left( 188 \left\lceil \log_2(2m) \right\rceil \cdot \left\lceil \sqrt{\frac{1}{\gamma}} \right\rceil \right)^{\frac{1}{2} \left\lceil \log_2(2m) \right\rceil \cdot \left\lceil \sqrt{\frac{1}{\gamma}} \right\rceil}$.
\end{enumerate}
\end{theorem}

\begin{theorem}[Polynomial approximation of intersection of halfspaces II]
\label{theorem:polynomial-approximation-2}
Let $v_1, v_2, \dots, v_m \in \R^d$ be vectors such that $\norm{v_1},
\norm{v_2}, \dots, \norm{v_m} \le 1$. Let $\gamma \in (0,1)$.
Define
$$
r = 2 \left\lceil \frac{1}{4} \log_2(4m + 1) \right\rceil + 1 \quad \text{and} \quad s = \left \lceil \log_2(1/\gamma) \right \rceil \; .
$$
Then, there exists a multivariate polynomial $p:\R^d \to \R$ such that
\begin{enumerate}
\item $\displaystyle p(x) \ge 1/2$ \\
for all $\displaystyle x \in R^+ = \bigcap_{i=1}^m \left\{ x \in \B(0,1) ~:~ \ip{v_i}{x} \ge \gamma \right\}$,

\item $\displaystyle p(x) \le - 1/2$ \\
for all $\displaystyle x \in R^- = \bigcup_{i=1}^m \left\{ x \in \B(0,1) ~:~ \ip{v_i}{x} \le - \gamma \right\}$,

@@

The proofs of the theorems can be found in
Appendix~\ref{section:proof-of-polynomial-approximation} in the supplementary
material. The geometric interpretation of the two regions described in parts 1
and 2 of the theorems is explained in Figure~\ref{figure:pizza-slice}. Similar
but weaker results were proved by~\citet{Klivans-Servedio-2008}. In particular,
our bounds in parts 1, 2, 3, 4 of
Theorems~\ref{theorem:polynomial-approximation-1}~and~\ref{theorem:polynomial-approximation-2}
are independent of the dimension $d$.

\begin{figure}
\begin{center}
\input{figures/pizza-slice}
\end{center}
\caption[]{The figure shows the two regions $R^+$ and $R^-$ used in parts 1 and
2 of
Theorems~\ref{theorem:polynomial-approximation-1}~and~\ref{theorem:polynomial-approximation-2}
for the case $m=d=2$ and a particular choice of vectors $v_1, v_2$ and margin
parameter $\gamma$. The separating hyperplanes $\ip{v_1}{x} = 0$ and
$\ip{v_2}{x} = 0$ are shown as dashed lines.}
\label{figure:pizza-slice}
\end{figure}

@@

Using the lemma and the polynomial approximation theorems, we can prove that the
mapping $\phi$ maps any set of weakly linearly separable examples to a strongly
linearly separable set of examples. Due to space constraints, we only sketch the
proof of Theorem~\ref{theorem:margin-transformation} here, and defer the full
proof to Appendix~\ref{section:full-proof-of-theorem-margin-transformation}.

\begin{proof}[Proof sketch of Theorem~\ref{theorem:margin-transformation}]
Since the examples $(x_1, y_1)$, $(x_2, y_2)$, $\dots$, $(x_T, y_T)$ are weakly
linearly separable with margin $\gamma$, there are vectors $w_1, w_2, \dots, w_K$
satisfying \eqref{equation:weak-linear-separability-1} and
\eqref{equation:weak-linear-separability-2}.

@@

\begin{align*}
\deg(p_i) & = \lceil \log_2(2K-2) \rceil \cdot \left\lceil \sqrt{\frac{2}{\gamma}} \right\rceil \; ,
\end{align*}
such that all examples $x$ in $R_i^+$ (resp. $R_i^-$) satisfy $p_i(x) \geq 1/2$
(resp. $p_i(x) \leq -1/2$).
Therefore, for all $t=1,2,\dots,T$, if $y_t = i$ then $p_i(x_t) \ge 1/2$,
 and if $y_t \neq i$ then $p_i(x_t) \le -1/2$, and
\begin{align*}
\norm{p_i} & \le \left(188 \lceil \log_2(2K-2) \rceil \cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil \right)^{\frac{1}{2} \lceil \log_2(2K-2) \rceil
\cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil} \; . \\
\end{align*}
By \autoref{lemma:norm-bound}, there exists $c_i \in \ell_2$ such that
$\ip{c_i}{\phi(x)}_{\ell_2} = p_i(x)$, and
\begin{align*}
\norm{c_i}_{\ell_2}
& \le \left(376 \lceil \log_2(2K-2) \rceil \cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil \right)^{\frac{1}{2} \lceil \log_2(2K-2) \rceil
\cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil} \; .
\end{align*}
Define vectors $u_i \in \ell_2$ as
\begin{align*}
u_i & = \frac{c_i}{\sqrt{K} \left(376 \lceil \log_2(2K-2) \rceil \cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil \right)^{\frac{1}{2} \lceil \log_2(2K-2) \rceil
\cdot \left \lceil \sqrt{\frac{2}{\gamma}} \right \rceil}} \; . \\
\end{align*}
